---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Loading and preprocessing the data
```{r loadFile}
# Reading the activity.csv file into a variable dataSet
dataSet<-read.csv("activity.csv")
# Loading the dplyr libary for easy data manipulation
library(plyr)
library(reshape2)
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(lattice)
```
## What is mean total number of steps taken per day?
```{r computeMeanTotalStepsPerDay}
# convert the date column of data set from type factor to Date
dataSet<-transform(dataSet,date=as.Date(dataSet$date))
# Proceeding to plot bar chart
dataSet%>%
group_by(date)%>%
summarise(mean=mean(steps,na.rm=TRUE)) %>%
ggplot(aes(date,mean))+geom_bar(stat="identity") + ggtitle("Mean total number of steps taken per day") + scale_x_date(breaks=unique(dataSet$date)[seq(from=0,to=61,by=15)])
dataSet%>%
    group_by(date)%>%
    summarise(mean=mean(steps,na.rm=TRUE),median=median(steps,na.rm=TRUE)) %>%
    kable()
```

## What is the average daily activity pattern?

```{r averageDailyPattern}
dataSet %>%
	group_by(interval) %>%
	summarise(mean=mean(steps,na.rm=TRUE)) %>%
	ggplot(aes(x=interval,y=mean),ymax=mean) + geom_line() 
```

```{r computeMaximumNumberOfSteps} 
maxVal<-subset(dataSet,dataSet$steps==max(dataSet$steps,na.rm=TRUE))
sprintf("The 5 minute interval %d contains the maximum number of %d steps", maxVal$interval, maxVal$steps)
```

## Inputing missing values
```{r computeSumOfNAs} 
sumOfNAs<-sum(is.na(dataSet))
sprintf("There are %d rows with missing values in the dataset", sumOfNAs)
```
### Proceeding to implement the following strategy to fill all of the missing values in the dataset
    * Identify the rows with missing values
    * Identify the mean values per interval for that row
    * Update the data set with the value
```{r substituteNAsWithMean}
    # Creating a new DataSet
    dataSetWithoutNAs<-dataSet
    # Function to calculate the mean value.
    # The input to the function is the dataSet and the interval for which mean has to be calculated
    meanPerInterval<-function(dataSet, intvl) mean(na.omit(subset(dataSet,dataSet$interval==intvl))$steps)
    # Identify the rows in the data set with missing values.
    nullRowIndex<-which(is.na(dataSetWithoutNAs))
    # Iterate through the rows
    for (i in nullRowIndex) {
        dataSetWithoutNAs[i,]$steps<-meanPerInterval(dataSetWithoutNAs,dataSetWithoutNAs[i,]$interval)
    }
```
```{r computeMeanTotalStepsPerDayWithoutNull}
dataSetWithoutNAs%>%
group_by(date)%>%
summarise(mean=mean(steps,na.rm=TRUE)) %>%
ggplot(aes(date,mean))+geom_bar(stat="identity") + ggtitle("Mean total number of steps taken per day")
dataSetWithoutNAs%>%
    group_by(date)%>%
    summarise(mean=mean(steps,na.rm=TRUE),median=median(steps,na.rm=TRUE)) %>%
    kable()
```

```{r computeMaximumNumberOfStepsWithoutNull} 
maxVal<-subset(dataSetWithoutNAs,dataSetWithoutNAs$steps==max(dataSetWithoutNAs$steps))
sprintf("The 5 minute interval %d contains the maximum number of %d steps", maxVal$interval, maxVal$steps)
```
## Are there differences in activity patterns between weekdays and weekends?
```{r checkActivityTypePerDay} 
# Create a new column Day and assign it values "Weekday" or "Weekend" depending on
# the value of date. Using weekday() function for this.
rownum<-nrow(dataSetWithoutNAs)
for(i in 1:rownum) {
	dayType=weekdays(as.Date(dataSetWithoutNAs[i,]$date))
	if(dayType=="Sunday" || dayType=="Saturday") {
		dataSetWithoutNAs[i,4]<-"Weekend"
	} else {
		dataSetWithoutNAs[i,4]<-"Weekday"
	}
}
names(dataSetWithoutNAs)[4]<-"Day"

dataSetDays <- ddply(dataSetWithoutNAs,
              .(interval, Day), 
              summarize, 
              mean = mean(steps, na.rm=TRUE))
xyplot(mean~interval|Day,data=dataSetDays,
       type='l',layout=c(1,2),xlab="Interval",ylab="Number of steps") 

```



